How Drug Usage During Labor and Delivery Affects Infant Health: A Predictive Analysis of Apgar5 Scores

### 2024-12-01 paper/paper.qmd
revise abstract and introduction

---
title: "How Drug Usage During Labor and Delivery Affects Infant Health: A Predictive Analysis of Apgar5 Scores"
subtitle: "Steroid Efficacy and Antibiotic Risks in Labor-Associated Care"
author: 
  - Yunkyung Ko
thanks: "Code and data are available at: [https://github.com/koyunkyung/infant_health](https://github.com/koyunkyung/infant_health)."
date: today
date-format: long
abstract: "How maternal and neonatal drug interventions during labor shape the health status of a newborn remains a key question in perinatal care. With the 2023 Natality Data for the United States, we examined the relationship between six labor-related treatments and Apgar5 scores, a standard measure of neonatal health. Using a Random Forest model for prediction and a Bayesian Linear Model for inference, we identified usage of steroids and chorioamnionitis as the most impactful, while antibiotics dosage showed minimal influence. The analysis shows significant disparities in treatment efficacy, and further suggests the need for optimized corticosteroid formulations or targeted antibiotic protocols."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
toc-title: "Table of Contents"
toc-depth: 2
toc-location: left
---

@@

The health of newborns is strongly influenced by the treatments administered during labor and delivery [@who_operational_guidance]. These interventions, including steroids, antibiotics, and anesthesia, are intended to address complications and improve neonatal outcomes, but their effectiveness varies widely [@anesthesiology_obstetric_anesthesia]. This paper uses the 2023 Natality Data for the United States [@rawdata] to analyze the impact of six specific treatments on neonatal health outcomes, measured using the Apgar5 score.

The estimand of interest is the average treatment effect (ATE) of six labor-associated interventions on neonatal health outcomes, as measured by Apgar5 scores. Specifically, we aim to quantify how the administration of each treatment changes the Apgar score compared to scenarios where the treatment is not applied [@tellingstories]. By applying predictive modeling via Random Forest and inferential analysis through Bayesian linear models, we assess not only the magnitude of these effects[@geeksforgeeks_random_forest] but also the underlying uncertainties[@strimmerlab_priors].

The analysis shows that prenatal exposure to steroids significantly improve Apgar scores (@tbl-rfsummary1), underscoring their role as a cornerstone of perinatal care. Conversely, antibiotics dosage doesn't show much contribution to the score distribution (@tbl-rfsummary1) and raises concerns about potential risks, including microbiota disruption [@narrow_spectrum_antibiotics]. Moreover, the analysis proposes significant disparities in the consistent application and effectiveness of these treatments (@fig-rfsumbar).

This study has its significance in that it states the need to improve health outcomes for mothers and their newborns globally. As recognized by the World Health Organization, a healthy start in life has significant repercussions for a person's health and well-being during infancy, childhood, and adulthood [@who_maternal_health]. Ensuring quality care in maternal and newborn health is integral to the right to health, equity, and the preservation of dignity for women and their babies [@who_maternal_health].

>>>>

---
title: "How Drug Usage During Labor and Delivery Affects Infant Health: A Predictive Analysis of Apgar5 Scores"
subtitle: "Steroid Efficacy and Antibiotic Risks in Labor-Associated Care"
author: 
  - Yunkyung Ko
thanks: "Code and data are available at: [https://github.com/koyunkyung/infant_health](https://github.com/koyunkyung/infant_health)."
date: today
date-format: long
abstract: "How maternal and neonatal drug interventions during labor shape the health status of a newborn remains a key question in perinatal care. With the 2023 Natality Data for the United States, the relationship between six labor-related treatments and Apgar5 scores, a standard measure of neonatal health, was examined. Using a Random Forest model for prediction and a Bayesian Linear Model for inference, we identified usage of steroids and chorioamnionitis as the most impactful, while antibiotics dosage showed minimal influence. The analysis shows disparities in treatment efficacy, and further suggests the need for optimized corticosteroid formulations or targeted antibiotic protocols."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
toc-title: "Table of Contents"
toc-depth: 2
toc-location: left
---

@@

Labor and birth is a time of excitement and anticipation, along with uncertainty and anxiety [@canada2023maternity]. This emotional complexity is often heightened by decisions surrounding medical interventions that directly impact maternal and newborn health [@reproductive2019labor]. Women frequently report anxiety in medication use especially during labor, driven by concerns about potential side effects [@reproductive2019labor] and insufficient information [@canada2023maternity]. To address these concerns, this paper analyzes the 2023 Natality Data for the United States [@rawdata], specifically evaluating the relationship between six labor-related treatments and neonatal health outcomes, as measured by the Apgar5 score [@natality2023].

The estimand is the average treatment effect (ATE) of six labor-associated interventions on neonatal health outcomes, quantified by Apgar5 scores [@rawdata]. Specifically, we aim to estimate how the administration of each treatment changes the Apgar score compared to scenarios where the treatment is not applied [@tellingstories]. By applying predictive modeling through a Random Forest model and inferential analysis through a Bayesian linear model, we assess not only the magnitude of these effects[@geeksforgeeks_random_forest] but also the underlying uncertainties[@strimmerlab_priors].

The analysis shows that prenatal exposure to steroids significantly improve Apgar scores (@tbl-rfsummary1). Conversely, antibiotics dosage doesn't show as much contribution to the score distribution (@tbl-rfsummary1) and rather raises concerns about potential risks, including microbiota disruption [@narrow_spectrum_antibiotics]. The paper therefore proposes significant disparities in the consistent application and effectiveness of these treatments (@fig-rfsumbar).

This study has its significance in that it states the need to improve health outcomes for mothers and their newborns. As recognized by the World Health Organization, a healthy start in life has significant repercussions for a person's health and well-being during infancy, childhood, and adulthood [@who_maternal_health]. Ensuring quality care in maternal and newborn health is integral to the right to health, equity, and the preservation of dignity for women and their babies [@who_maternal_health].
